---
title: "Modeling Price Change of Seasonal Fruits-Flavored Food Products: A Predictive Analysis"
subtitle: "The Price of Strawberry and Banana-Flavored Products will Increase after November, with Strawberry Flavored Products Rising More in Price"
author: 
  - Yiyi Feng
thanks: "Code and data are available at: https://github.com/kqlqkqlqF/Insights-and-Predictions-for-the-U.S.-Election.git."
date: today
date-format: long
abstract: "This study develops a model to predict price changes for seasonal fruit-flavored products during in-season and off-season periods. It forecasts price increases for strawberry and banana-flavored products after November, with a larger rise for strawberry. Vendor and month are key factors in price trends, but the complexity of price changes calls for further investigation. Due to the complex dynamics of seasonal fruit-flavored product price changes, improved data collection is essential for more accurate forecasting."
format:
  pdf:
    toc: true
    number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

#### Preamble ####
# Purpose: Help constructing the "Modeling Price Change of Seasonal Fruits-Flavored Food Products: A Predictive Analysis" paper
# Author: Yiyi Feng
# Date: 29 November 2024
# Contact: yiyi.feng@mail.utoronto.ca

####Workspace setup ####

library(dplyr)
library(tibble)
library(here)
library(modelsummary)
library(tidyverse)
library(arrow)
library(ggplot2)
library(janitor)
library(knitr)
library(kableExtra)
library(caret)
library(randomForest)
library(rstanarm)

model_data <- read_parquet(here::here("data/02-analysis_data/combined_model_data.parquet"))
rain_data <- read_parquet(here::here("data/02-analysis_data/average_rain_data.parquet"))

```


# Introduction

As a fruit enthusiast, I’ve noticed seasonal fruits show significant price changes between in-season and off-season periods. This raises the question: do seasonal fruit-flavored products follow similar patterns, and what drives these changes? What are the key factors driving these price changes? This study uses Canadian grocery price data from eight major vendors to build a predictive model analyzing seasonal fruit-flavored product pricing and identifying key factors influencing price dynamics.

Our estimand focuses on the price changes in seasonal fruit-flavored products during in-season and off-season periods, focusing on strawberries and bananas due to their popularity as summer and winter fruits []. These choices ensure a sufficiently large sample size for analysis. Other seasonal fruits, such as watermelons, were excluded due to their relatively smaller consumer base in North America compared to Asia.

We used a linear model to estimate price changes for strawberry- and banana-flavored products, including predictors like vendor, month, rainfall, food category, and their interactions. The goal was to understand how these factors affect price fluctuations and seasonal pricing patterns.

Our model predicts small price increases for both banana and strawberry-flavored products, with average monthly increases of 0.0085 and 0.0103, respectively. Monthly trends show stability in July, slight decreases in August and September, and increases in October and November. The most significant factors influencing price changes are vendor, month, and the interaction between month and food category, emphasizing the important roles of vendors, timing, and product category in price fluctuations.

The remainder of this paper is structured as follows: [@sec-data] provides an overview of the dataset, details of the parameters, outcome and predictor variables, and the packages used during processing. [@sec-model] explains the modeling approach, and best model selection, justifying the choice of predictors and outlining the methods used to forecast the price change of banana and strawberry-flavored products. [@sec-results] presents the findings, including a summary of the predicted price change for each flavored product, and figures demonstrating the price change distribution across month and food categories. In [@sec-discussion], we discuss the implications of these results, the limitations of our analysis, and what can we do next to improve our model. Additional methodological details and diagnostics are included in the appendix.


# Data {#sec-data}

## Overview

We used R [@citeR] to analyze Canadian grocery price data. The dataset, from Jacob Filipp's Project Hammer [@hammer], tracks price changes from eight vendors: Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods, from February 28, 2024 to November 26, 2024. To examine factors influencing the price of seasonal fruit-flavored foods, we included data on vendor, rainfall, month, and food category, along with rainfall data from the "About Rain Gauge Locations and Precipitation" dataset for opendataToronto [@open].

Several R packages were essential for our data manipulation, modeling, and visualization efforts. dplyr facilitated data transformation and summarization [@dplyr], while modelsummary enhanced the presentation of model outputs [@modelsummary]. kableExtra improved data presentation with customizable tables [@kableExtra], and testthat ensured code reliability [@testthat]. Lubridate and stringr simplified date-time handling and string manipulation [@lubridate] [@stringr]. For efficient data storage, tibble and arrow enhanced our workflow with memory-efficient data frames and fast data reading/writing [@tibble] [@arrow]. Caret and randomForest supported machine learning, and rstanarm provided Bayesian regression tools [@caret] [@randomForest] [@rstanarm]. Finally, ggplot2 powered visualizations, and knitr streamlined reproducible reporting [@ggplot2] [@knitr]. This robust package ecosystem aligned with best practices, as noted in [@tellingstories].

## Measurement

This section outlines how we transformed the raw Canadian grocery price data into a structured dataset for analysis. We focused on seasonal fruit-flavored foods, selecting bananas and strawberries as representatives. Bananas were chosen for their popularity in winter, and strawberries for spring. These fruits have a larger consumer base in Canada compared to other seasonal fruits like watermelon and pomegranate, making data collection for model building more feasible.

The original Canadian grocery price data was collected by Jacob Filipp through screen-scraping the website interfaces of eight vendors and compiled into the \textbf{Project Hammer} dataset. This dataset contains two tables: \textbf{Hammer 4 Raw} and \textbf{Hammer 4 Product}.
\textbf{Hammer 4 Raw} includes the scraping time, product name, single-item price, unit price (e.g., price per 100g), past prices, additional information (e.g., availability or discounts), and a unique product ID.
\textbf{Hammer 4 Product} provides detailed information, such as brand, vendor, sales unit, and product detail page links.
Since the data comes from website scraping, it contains many gaps and ambiguities, making cleaning difficult. Details about data collection and cleaning are provided in Appendix A.

We considered sources such as Statistics Canada, the Retail Council of Canada, and Open Data Portals, but after reviewing their datasets, we found they did not meet our study’s requirements. These datasets focus on broad food categories, not specific flavored products like strawberries or bananas. Additionally, they lack the detailed time scale and vendor-level data necessary for tracking monthly price changes, which are crucial for modeling seasonal price shifts. As a result, we determined these sources were not suitable for our research.

To better analyze price changes for banana and strawberry-flavored foods, we added data from the \textbf{"About Rain Gauge Locations and Precipitation"} dataset from Open Data Toronto []. This dataset includes rain gauge locations across Toronto and recorded precipitation. Rainfall was included as a predictor since consumer demand for seasonal fruits often correlates with weather conditions, such as a preference for watermelon and strawberries during dry summers []. The dataset only uses rainfall data because the original grocery price data lacks sales location details.

This structured dataset enables us to analyze trends in seasonal fruit-flavored food prices over time. The goal is to examine price trends for these foods during in-season and off-season sales and identify key influencing factors.


## Outcome Variables

### Change of Monthly Averaged Price for Banana and Strawberry Flavored Product

[@fig-one] shows the monthly average price distribution of banana and strawberry-flavored products. Most products are priced between 0 and 10 dollars, with few exceeding this range and prices above 20 dollars being rare. This indicates that the majority of sales fall within the lower price range. The price distributions of both flavors are similar, though strawberry-flavored products have a slightly smaller share in the 0 to 10-dollar range compared to banana-flavored products.

```{r}
#| label: fig-one
#| fig-cap: Distribution of Monthly Averaged Price for Banana and Strawberry Flavored Product
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Create the plot for average price distribution
ggplot(model_data, aes(x = monthly_avg_price)) +
  geom_histogram(aes(y = ..density..),
                 bins = 30,
                 fill = "blue",
                 color = "black",
                 alpha = 0.7
  ) +
  geom_density(color = "black", fill = "salmon", alpha = 0.6) +
  labs(
    x = "Average Price",
    y = "Density"
  ) +
  theme_minimal() +
  facet_wrap(~flavor) +
  theme(
    strip.text = element_text(size = 12), # Adjust facet label size
    axis.title = element_text(size = 14), # Adjust axis title size
    axis.text = element_text(size = 12)  # Adjust axis label size
  )


```

```{r}
#| label: fig-two
#| fig-cap: Distribution of Monthly Averaged Price Change for Banana and Strawberry Flavored Product from July 2024 to Novermber 2024
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Create the plot for price distribution by month
ggplot(model_data, aes(x = as.factor(month), y = price_change, color = price_change)) +
  geom_violin(fill = "blue", alpha = 0.4) +  # Use violin plot to show the distribution of the average prices
  geom_jitter(width = 0.1, alpha = 0.7) +  # Scatter plot for individual data points, jittered for visibility
  scale_color_gradient(
    low = "blue",
    high = "red",
    name = "Price Change"  # Change the legend title
  ) +
  labs(
    x = "Month",
    y = "Average Price Change"
  ) +
  theme_minimal() +
  facet_wrap(~flavor) +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    strip.text = element_text(size = 12)
  )


```

[@fig-two] shows the monthly price changes for banana and strawberry-flavored products. Strawberry-flavored products exhibit greater fluctuations, ranging within ±2.5 dollars, compared to banana-flavored products, which vary within ±1.5 dollars. Neither flavor shows a clear overall trend of increasing or decreasing prices across the months.

This outcome challenges our assumption. We expected banana-flavored product prices to rise between July and November due to their popularity in fall and winter and strawberry-flavored prices to decrease during this period as a summer flavor.


## Predictor Variables

### Summary of Predictor Variables

- **Vendor:** Includes Canada’s eight major suppliers: Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods.


- **Average Rainfall Per Month:** Monthly average rainfall for Toronto from July to November 2024, measured in millimeters, based on the "About Rain Gauge Locations and Precipitation" dataset.

- **Food Category:** Banana and strawberry-flavored products are divided into five categories: beverage, yogurt, flavored tea, solid snack, and fruit. "Beverage" includes all drinks except tea and yogurt-based beverages, "flavored tea" covers tea drinks and flavored tea bags, and "fruit" includes various forms of banana or strawberry fruit.

- **Month:** Indicates the month when the price data for each product was collected, covering July to November.

### Vendor

Based on [@fig-vendor], strawberry-flavored products are significantly more common than banana-flavored ones, with the dataset showing 3-4 times as many strawberry-flavored items. This suggests that strawberry flavors appeal to a larger audience. Save-On-Foods has the smallest share for both flavors, which may reflect a narrower focus on fruit-flavored products or a smaller overall selection of food types.

```{r}
#| label: fig-vendor
#| fig-cap: Distribution of the Counts of Banana and Strawberry Flavored Food Offered in Each Vendor
#| echo: false
#| warning: false
#| message: false

# Create a bar plot for vendor distribution
ggplot(model_data, aes(x = vendor)) +
  geom_bar(fill = "blue", color = "black", alpha = 0.7) + 
  labs(
    x = "Vendor",
    y = "Count"
  ) +
  theme_minimal() +
  facet_wrap(~flavor) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels for better readability
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )


```

### Average Rainfall Per Month

[@fig-rain] displays the average monthly rainfall in Toronto, measured in millimeters. Rainfall decreases steadily from July to October, with October having the lowest levels, followed by a rebound in November. Rainfall was included as a predictor due to its strong correlation with time and seasonal changes, making it a potentially valuable factor in the model.

```{r}
#| label: fig-rain
#| fig-cap: Average Rainfall in Toronto from July 2024 to November 2024
#| echo: false
#| warning: false
#| message: false

# Assuming rain_data is your dataset
ggplot(rain_data, aes(x = as.factor(month), y = avg_rainfall)) +
  geom_bar(stat = "identity", fill = "blue", color = "black", alpha = 0.7) + 
  labs(
    x = "Month",
    y = "Average Rainfall (mm)",
    title = "Average Rainfall per Month"
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )


```

### Food Category

In [@fig-category], we can see that solid snacks have the highest overall prices, while flavored tea has the lowest overall prices. Additionally, the price distribution for solid snacks, fruit, and beverages is relatively dispersed, whereas yogurt and flavored tea show more tightly clustered price distributions. We believe this is because flavored tea and yogurt are more concentrated categories compared to the others.

```{r}
#| label: fig-category
#| fig-cap: Distribution of Monthly Averaged Price for Banana and Strawberry Flavored Product within Category
#| echo: false
#| warning: false
#| message: false
# Convert pct column to numeric if it is not already

# Create the plot for price distribution by category
ggplot(model_data, aes(x = category, y = monthly_avg_price, color = monthly_avg_price)) +
  geom_violin(fill = "blue", alpha = 0.4) +  # Use violin plot to show the distribution of the average prices
  geom_jitter(width = 0.1, alpha = 0.7) +  # Scatter plot for individual data points, jittered for visibility
  scale_color_gradient(
    low = "blue",
    high = "red",
    name = "Average Price"  # Change the legend title
  ) +
  labs(
    x = "Month",
    y = "Average Price"
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    strip.text = element_text(size = 12)
  )


```

### Month

[@fig-month] shows the distribution of the Monthly average price for Banana and strawberry-flavored products over time. It is clear that strawberry-flavored products are priced higher than banana-flavored products in all months. However, neither flavor shows a noticeable increasing or decreasing trend in price distribution across the months.

```{r}
#| label: fig-month
#| fig-cap: Distribution of Monthly Averaged Price for Banana and Strawberry Flavored Product from July 2024 to Novermber 2024
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Create the plot for price distribution by month
ggplot(model_data, aes(x = as.factor(month), y = monthly_avg_price, color = monthly_avg_price)) +
  geom_violin(fill = "blue", alpha = 0.4) +  # Use violin plot to show the distribution of the average prices
  geom_jitter(width = 0.1, alpha = 0.7) +  # Scatter plot for individual data points, jittered for visibility
  scale_color_gradient(
    low = "blue",
    high = "red",
    name = "Average Price"  # Change the legend title
  ) +
  labs(
    x = "Month",
    y = "Average Price"
  ) +
  theme_minimal() +
  facet_wrap(~flavor) +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    strip.text = element_text(size = 12)
  )

```


# Model {#sec-model}

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

model_linear_1 <-
  readRDS(file = here::here("models/model_linear_1.rds"))

model_linear_2 <-
  readRDS(file = here::here("models/model_linear_2.rds"))

model_linear_3 <-
  readRDS(file = here::here("models/model_linear_3.rds"))

model_linear_4 <-
  readRDS(file = here::here("models/model_linear_4.rds"))

model_linear_5 <-
  readRDS(file = here::here("models/model_linear_5.rds"))

```

The goal of this section is to build a predictive model for price changes in banana and strawberry-flavored products. The main challenge is to create a model that captures price fluctuations accurately, despite the limited sample size and temporal scope of the data. To address this, we evaluated different model specifications to identify the best approach for achieving reliable predictions.

We decided to model the monthly average unit price changes for banana and strawberry-flavored products, instead of predicting more specific metrics like price per 100g. This choice was made due to limitations in the dataset, which includes various measurement units (e.g., per 100g, per 100 ml, per kg, per lb) and the fact that many products lack detailed measurement information. While this approach may slightly reduce the model’s performance, it allowed us to focus on unit prices.

We included several predictors in the model, such as Vendor, Food Category, Average Rainfall, and Month, and considered potential interactions among these variables. For example, rainfall and month are often closely correlated, and food categories might interact with both month and vendor. We developed a series of linear models and systematically compared them, testing different combinations of interaction terms to identify the model that best predicts price changes.

## Model Set-up

Our goal is to model the price changes of strawberry and banana-flavored products using Vendor, Food Category, Average Rainfall, and Month as predictors. The model includes interaction terms to capture how combinations of these factors influence price changes, offering insights into the key factors that drive the pricing of seasonal fruit-flavored products.

\
\begin{align*}
\mathrm{Pct}_i = &\ \beta_0 + \beta_1 \cdot \mathrm{Vendor}_i + \beta_2 \cdot \mathrm{Category}_i + \beta_3 \cdot \mathrm{Month}_i + \beta_4 \cdot \mathrm{Average Rainfall}_i \\
& + \beta_5 \cdot (\mathrm{Vendor}_i \times \mathrm{Category}_i) + \beta_6 \cdot (\mathrm{Category}_i \times \mathrm{Month}_i) \\
& + \beta_7 \cdot (\mathrm{Month}_i \times \mathrm{Average Rainfall}_i) + \beta_8 \cdot (\mathrm{Category}_i \times \mathrm{Average Rainfall}_i) + \epsilon_i
\end{align*}
\

Where

-   $y_i$ : The percentage of support for candidate in poll i.
-   $\beta_0$: Intercept term, representing the predicted `Average price of the Product` when all independent variables are 0.
-   $\beta_1$: Main effect of `Vendor`, capturing the influence of the candidate.
-   $\beta_2$: Main effect of `Category`, reflecting the influence of how recent the poll is on `Average price of the Product`.
-   $\beta_3$: Main effect of `Month`, indicating the impact of different states on `Average price of the Product`.
-   $\beta_4$: Main effect of `Average Rainfall`, indicating the impact of different states on `Average price of the Product`.
-   $\beta_5$: Interaction effect between `Vendor` and `Category`, capturing the combined influence of the candidate and state.
-   $\beta_6$: Interaction effect between `Category` and `Month`, reflecting the joint impact of recency and state on `pct`.
-   $\beta_7$: Interaction effect between `Month` and `Average rainfall`, representing the combined influence of the candidate and recency of the poll.
-   $\beta_8$: Interaction effect between `Category` and `Average rainfall`, representing the combined influence of the candidate and recency of the poll.
-   $\epsilon_i$: The error term, assumed to follow a normal distribution with mean 0.

### Model Interpretation


This regression model predicts price changes for strawberry and banana-flavored products by incorporating both main effects and interaction terms. The intercept represents the baseline price change when all predictors are zero. Main effects, such as vendor, average rainfall, food category, and month, capture the independent influence of each factor on price. For example, strawberry-flavored products, typically associated with summer, may see price drops as the months move into winter.

To capture more complex relationships, the model includes two-way interactions, such as vendor and category, category and month, and month and average rainfall. These interactions help explain how factors influence each other. For instance, some vendors may focus on specific products, like beverages, while the category-month interaction reflects seasonal demand shifts, such as reduced beverage demand in Canada’s colder months. The rainfall-month interaction shows their direct relationship, with June being the wettest month and November the driest in Canada.

An error term is included to account for unexplained variability, improving the model's predictive accuracy. By considering both individual and interactive effects, the model provides a comprehensive approach to forecasting price changes for these products.

## Model Justification

```{r}
#| label: tbl-linear
#| tbl-cap: "Predicted Average Supporting Percentages for Donald Trump vs. Kamala Harris by Random Forest"
#| echo: false
#| warning: false
#| message: false

# Extract summary statistics and add variables included in each model
model_summary <- tibble(
  Model = c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5"),
  Variables = c(
    linebreak("Vendor, Category, Month, Average Rainfall"), 
    linebreak("Vendor, Category × Month, Average Rainfall"), 
    linebreak("Vendor, Category, Month × Average Rainfall"),
    linebreak("Vendor, Month, Category × Average Rainfall"),
    linebreak("Vendor × Category, Month, Average Rainfall")
  ),
  `R²` = round(c(summary(model_linear_1)$r.squared, 
                 summary(model_linear_2)$r.squared, 
                 summary(model_linear_3)$r.squared,
                 summary(model_linear_4)$r.squared,
                 summary(model_linear_5)$r.squared), 5),
  `Adjusted R²` = round(c(summary(model_linear_1)$adj.r.squared,
                          summary(model_linear_2)$adj.r.squared, 
                          summary(model_linear_3)$adj.r.squared,
                          summary(model_linear_4)$adj.r.squared,
                          summary(model_linear_5)$adj.r.squared), 5),
  `AIC` = round(c(AIC(model_linear_1), AIC(model_linear_2), AIC(model_linear_3), AIC(model_linear_4), AIC(model_linear_5)), 5),
  `BIC` = round(c(BIC(model_linear_1), BIC(model_linear_2), BIC(model_linear_3), BIC(model_linear_4), BIC(model_linear_5)), 5),
  `RMSE` = round(c(sqrt(mean(residuals(model_linear_1)^2)), 
                   sqrt(mean(residuals(model_linear_2)^2)), 
                   sqrt(mean(residuals(model_linear_3)^2)),
                   sqrt(mean(residuals(model_linear_4)^2)),
                   sqrt(mean(residuals(model_linear_5)^2))), 5)
)

# Display the table using kable
model_summary %>%
  kable(digits = 4, escape = FALSE) %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(2, width = "8em")


```

[@tbl-linear] summarizes the performance metrics for five models, each with different interactions, while the first model contains no interaction.

Model 1, which includes basic predictors (Vendor, Category, Month, and Average Rainfall), exhibits poor predictive performance with an R² of 0.0038 and an RMSE of 0.5098, indicating limited explanatory power. Model 2 improves slightly by incorporating an interaction between Category and Month, increasing R² to 0.00890 and lowering RMSE to 0.5085, suggesting minor gains in prediction accuracy. Model 3, which includes an interaction between Month and Average Rainfall, yields no improvement over Model 1, with identical R² (0.00383) and RMSE (0.5098), highlighting the limited value of this interaction.

Model 4 adds an interaction between Category and Average Rainfall, resulting in a marginally higher R² of 0.0043 but with negligible impact on RMSE (0.5097). Finally, Model 5 introduces an interaction between Vendor and Category, leading to an R² of 0.00821 and an RMSE of 0.5087, showing slight improvement but still limited overall explanatory power.

In conclusion, none of the models provide strong predictive accuracy, but Model 2 achieves the best balance among them with the highest R² (0.0089) and lowest RMSE (0.50853). However, the overall low R² values indicate that these predictors and interactions capture only a small portion of the variation in price changes.

## Optimization

```{r}

tree_data <- model_data

tree_data$month_category <- with(tree_data, interaction(month, category))

interaction_terms <- model.matrix(~month * category - 1, data=tree_data)
tree_data <- cbind(tree_data, interaction_terms)
main_effects <- tree_data[c("vendor", "avg_rainfall")]
X <- cbind(main_effects, interaction_terms)
y <- tree_data$price_change

set.seed(333)
trainIndex <- createDataPartition(y, p = 0.8, list = FALSE)
train_data <- X[trainIndex,]
test_data <- X[-trainIndex,]
train_y <- y[trainIndex]
test_y <- y[-trainIndex]

tuneGrid <- expand.grid(mtry = c(2, 5, 10))

set.seed(333)
rf_model <- train(
  x = train_data,
  y = train_y,
  method = "rf",
  tuneGrid = tuneGrid,
  trControl = trainControl(method = "cv", number = 10),
  metric = "RMSE",
  ntree = 200
)

```

```{r}
#| label: tbl-rfsum
#| echo: false
#| warning: false
#| message: false

# Extract RMSE and other metrics from the model
model_metrics <- tibble(
  Metric = c("RMSE", "Number of Trees", "Best mtry"),
  Value = c(
    min(rf_model$results$RMSE),  # Minimum RMSE from cross-validation
    rf_model$finalModel$ntree,  # Number of trees used
    rf_model$bestTune$mtry      # Best mtry value selected
  )
)

# Display the model metrics table
model_metrics %>%
  kable(digits = 4, escape = FALSE) %>%
  kable_styling(full_width = FALSE) %>%
  add_header_above(c(" " = 1, "Model Evaluation Metrics" = 1)) %>%
  column_spec(1, bold = TRUE, width = "12em")

# Extract variable importance
var_importance <- varImp(rf_model, scale = TRUE)

# Convert to tibble for display
importance_table <- as_tibble(var_importance$importance, rownames = "Variable") %>%
  arrange(desc(Overall)) %>%  # Sort by importance
  mutate(Overall = round(Overall, 2))  # Round importance scores

# Summing up rows for month6 to month11
month_sum <- importance_table %>%
  filter(grepl("^month[6-9]|month1[01]$", Variable)) %>% # Select month6 to month11
  summarise(Overall = sum(Overall)) %>%                # Sum their importance
  mutate(Variable = "month")                           # Add the new row label

# Removing rows for month6 to month11 and adding the summarized row
importance_table_transformed <- importance_table %>%
  filter(!grepl("^month[6-9]|month1[01]$", Variable)) %>% # Remove month6 to month11 rows
  bind_rows(month_sum)                                    # Add the new row


# Summing up rows for categoryFruit, categorySolid, and categoryYogurt
category_sum <- importance_table_transformed %>%
  filter(Variable %in% c("categoryFruit", "categorySolid", "categoryYogurt")) %>% # Select relevant rows
  summarise(Overall = sum(Overall)) %>%                                         # Sum their importance
  mutate(Variable = "category")                                                 # Add the new row label

# Removing rows for categoryFruit, categorySolid, and categoryYogurt
# Adding the summarized row
importance_table_transformed_2 <- importance_table_transformed %>%
  filter(!Variable %in% c("categoryFruit", "categorySolid", "categoryYogurt")) %>% # Remove old rows
  bind_rows(category_sum) 


# Summing up rows for all remaining variables (excluding Category, month, vendor, avg_rainfall)
month_category_sum <- importance_table_transformed_2  %>%
  filter(!Variable %in% c("category", "month", "vendor", "avg_rainfall")) %>% # Select remaining rows
  summarise(Overall = sum(Overall)) %>%                                       # Sum their importance
  mutate(Variable = "month:category")                                         # Add the new row label

# Removing old rows and adding the new row
final_importance_table <- importance_table_transformed_2  %>%
  filter(Variable %in% c("category", "month", "vendor", "avg_rainfall")) %>% # Keep important rows
  bind_rows(month_category_sum)                                              # Add the new summarized row

# Arrange the table for better readability
final_importance_table <- final_importance_table %>%
  arrange(desc(Overall))

# Rename avg_rainfall to average rainfall
final_importance_table <- final_importance_table %>%
  mutate(Variable = ifelse(Variable == "avg_rainfall", "average rainfall", Variable))


# Display variable importance table
final_importance_table  %>%
  kable(digits = 2, escape = FALSE) %>%
  kable_styling(full_width = FALSE) %>%
  add_header_above(c(" " = 1, "Variable Importance" = 1)) %>%
  column_spec(1, bold = TRUE, width = "12em")


```

Due to the limitations of linear models in handling interaction terms, we propose an optimized model—Random Forest. The consistently poor performance of linear models suggests that a linear relationship may not adequately explain the price changes of banana and strawberry flavored products in relation to the predictors. We chose Random Forest because our data includes interaction terms, and linear analysis is not suitable for such data. Random Forest excels at handling interactions and nonlinear relationships, as it can automatically identify and leverage these complex interactions through its decision tree structure without the need for manual adjustments. In contrast, linear models have limited capacity to handle interactions, requiring manual specification, and are less effective at capturing nonlinear effects.

[@tbl-rfsum] presents the prediction results obtained using the Random Forest model.

Compared to the best performing Linear Model 2, the Random Forest model demonstrates higher predictive accuracy. Model 2, which includes basic predictors (Vendor, Category, Month, and Average Rainfall) along with an interaction between Category and Month, achieves an RMSE of 0.50853, indicating limited improvement in prediction accuracy. On the other hand, the Random Forest model, with 200 trees, achieves an RMSE of 0.5076, showcasing a slight reduction in error and an overall better fit to the data.

The Random Forest model further outperforms the linear model in handling complex relationships and interactions. By incorporating important variables such as Vendor, Month, Month:Category, and Average Rainfall, and automatically detecting complex interactions between predictors, the Random Forest model demonstrates its ability to account for these factors effectively. Notably, the model's variable importance analysis reveals that 'Vendor' has the most significant impact on predictions, followed by 'Month' and 'Month:Category', highlighting the model’s capacity to prioritize key features. In conclusion, the Random Forest model provides stronger predictive power and better captures the underlying complexities of the data, surpassing the linear model in performance.

# Result {#sec-results}


```{r}
#| label: tbl-resultone
#| tbl-cap: "Summary Statistics of Predicted Change in Average Price of Strawberry and Banana Flavored Product by Linear Model Two"
#| message: false
#| echo: false
#| warning: false

# Step 1: Predict price changes for the existing dataset
predict_data_lm <- model_data %>%
  mutate(Predicted_Price_Change = predict(model_linear_2, newdata = model_data))

# Step 2: Filter the dataset for 'strawberry' and 'banana' flavor products
flavor_data <- predict_data_lm %>%
  filter(flavor %in% c("strawberry", "banana"))

# Step 3: Calculate average predicted price change for each flavor across the dataset
flavor_summary <- flavor_data %>%
  group_by(flavor) %>%
  summarise(
    Avg_Price_Change = mean(Predicted_Price_Change, na.rm = TRUE),
    Median_Price_Change = median(Predicted_Price_Change, na.rm = TRUE),
    Min_Price_Change = min(Predicted_Price_Change, na.rm = TRUE),
    Max_Price_Change = max(Predicted_Price_Change, na.rm = TRUE),
    SD_Price_Change = sd(Predicted_Price_Change, na.rm = TRUE),
    Total_Products = n()
  )

# Step 4: Present the summary using kable
flavor_summary %>%
  kable(
    col.names = c("Flavor", "Avg Change", "Median Change", "Min Change", "Max Change", "SD of Change", "Sample Size"),
    digits = 4,
    booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "scale_down"))



```

[@tbl-resultone] summarizes the price change statistics for banana and strawberry flavored products predicted by linear model two.

Banana flavored products have an average price change of 0.0091, with a median change of 0.0181 across 1,376 products. Strawberry-flavored products show a slightly higher average change of 0.0100, with a median of 0.0207. For both of the flavors, their price changes  range from -0.109 to 0.1125. However, for banana flavored product, their sample size is roughly only a quarter of strawberry product sample size.Overall, strawberry-flavored products exhibit slightly higher average price changes and more variation compared to banana-flavored products.

```{r}
#| label: tbl-resulttwo
#| tbl-cap: "Summary Statistics of Predicted Change in Average Price of Strawberry and Banana Flavored Product by Random Forest"
#| echo: false
#| warning: false
#| message: false


tree_data <- tree_data[, !duplicated(names(tree_data))]

tree_data <- tree_data %>%
  mutate(PredictedPriceChange = predict(rf_model, newdata = tree_data))

results <- tree_data %>%
  group_by(flavor) %>%
  summarise(AveragePredictedPriceChange = mean(PredictedPriceChange))

total_change <- sum(results$AveragePredictedPriceChange)

results %>%
  kable(
    col.names = c("Flavor", "Average Predicted Price Change"),
    digits = 4,
    booktabs = TRUE
  ) %>%
  kable_styling(latex_options = c("striped", "scale_down"))

```

[@tbl-resulttwo] summarizes the price change statistics for banana and strawberry flavored products predicted by random forest model.

Banana flavored products have an average price change of 0.0085, while strawberry flavored products show a slightly higher average change of 0.0103. This result means that for random forest estimation, the price of both banana and strawberry flavored products are gonna increase, but only by very little scale.


```{r}
#| label: fig-lmone
#| fig-cap: Distribution of Linear Model Two Predicted Monthly Price Changes for Banana and Strawberry Products by Month  
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Create the plot for price distribution by month
ggplot(predict_data_lm, aes(x = as.factor(month), y = Predicted_Price_Change, color = Predicted_Price_Change)) +
  geom_violin(fill = "blue", alpha = 0.4) +  # Use violin plot to show the distribution of the average prices
  geom_jitter(width = 0.1, alpha = 0.7) +  # Scatter plot for individual data points, jittered for visibility
  scale_color_gradient(low = "blue", high = "red") +  # Color gradient based on the price
  labs(
    x = "Month",
    y = "Average Price") +
  theme_minimal() +
  facet_wrap(~flavor) +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    strip.text = element_text(size = 12)
  ) 

```

[fig-lmone] illustrates the predicted monthly price changes for strawberry and banana-flavored products. The price trends show no significant differences between the two flavors, with nearly stable prices in July, a slight dip in August, and minor increases in October and November. October saw the highest increase, while August experienced the largest decrease. However, the average price changes remained within ±0.1 units, indicating limited fluctuation.

```{r}
#| label: fig-lmtwo
#| fig-cap: Distribution of Linear Model Two Predicted Monthly Price Changes for Banana and Strawberry Products by Category
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Create the plot for price distribution by month
ggplot(predict_data_lm, aes(x = category, y = Predicted_Price_Change, color = Predicted_Price_Change)) +
  geom_violin(fill = "blue", alpha = 0.4) +  # Use violin plot to show the distribution of the average prices
  geom_jitter(width = 0.1, alpha = 0.7) +  # Scatter plot for individual data points, jittered for visibility
  scale_color_gradient(low = "blue", high = "red") +  # Color gradient based on the price
  labs(
    x = "Food Category",
    y = "Average Price"
  ) +
  theme_minimal() +
  facet_wrap(~flavor) +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    strip.text = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels
  )

```
[fig-lmtwo], like [fig-lmone], uses predictions from Linear Model 2. However, while [fig-lmone] shows predicted price changes across months, [fig-lmtwo] presents their distribution across food categories. Similar to the monthly distribution, there is no significant difference between strawberry and banana-flavored products across categories, except for flavored tea. Banana-flavored tea shows a downward price trend with fewer products compared to strawberry-flavored tea, which exhibits rising prices. For beverage, fruit, and solid snack categories, both flavors demonstrate increasing price trends, while yogurt prices remain largely stable.

```{r}
#| label: fig-rfone
#| fig-cap: Distribution of Random Forest Model Predicted Monthly Price Changes for Banana and Strawberry Products by Month
#| echo: false
#| eval: true
#| warning: false
#| message: false

predict_data_rf <- model_data %>%
  mutate(predicted_price_change = predict(model_rf, newdata = model_data))

# Create the plot for price distribution by month
ggplot(predict_data_rf, aes(x = as.factor(month), y = predicted_price_change, color = predicted_price_change)) +
  geom_violin(fill = "blue", alpha = 0.4) +  # Use violin plot to show the distribution of the average prices
  geom_jitter(width = 0.1, alpha = 0.7) +  # Scatter plot for individual data points, jittered for visibility
  scale_color_gradient(low = "blue", high = "red") +  # Color gradient based on the price
  labs(
    x = "Month",
    y = "Average Price") +
  theme_minimal() +
  facet_wrap(~flavor) +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    strip.text = element_text(size = 12)
  )


```

To address the limitations of the linear model, we also generated predictions using the random forest model, as shown in [@fig-rfone] and [@fig-rftwo].

[@fig-rfone] illustrates the predicted price changes for banana and strawberry-flavored products across months. While the differences are not pronounced, there are notable variations in their distributions. In July, August, and September, prices for both flavors decline, with banana-flavored products experiencing a steeper drop. By October, prices for both flavors rise, with a sharper increase for banana-flavored products. In November, price increases continue but at a slower pace, with some products, particularly banana-flavored ones, showing significant price reductions. Overall, the price changes largely fall within a range of ±0.1 units.

```{r}
#| label: fig-rftwo
#| fig-cap: This figure shows the distribution of monthly price changes for banana and strawberry products. A higher concentration around zero indicates more stable prices, while wider distributions suggest greater variability.
#| echo: false
#| eval: true
#| warning: false
#| message: false

predict_data_rf <- model_data %>%
  mutate(predicted_price_change = predict(model_rf, newdata = model_data))

# Create the plot for price distribution by month
ggplot(predict_data_rf, aes(x = category, y = predicted_price_change, color = predicted_price_change)) +
  geom_violin(fill = "blue", alpha = 0.4) +  # Use violin plot to show the distribution of the average prices
  geom_jitter(width = 0.1, alpha = 0.7) +  # Scatter plot for individual data points, jittered for visibility
  scale_color_gradient(low = "blue", high = "red") +  # Color gradient based on the price
  labs(
    x = "Month",
    y = "Average Price",
    title = "Average Price Distribution by Month"
  ) +
  theme_minimal() +
  facet_wrap(~flavor) +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    strip.text = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels
  )


```

[@fig-rftwo] presents the predicted price changes for banana and strawberry-flavored products across food categories. Notable differences emerge in their distribution across categories. Prices for beverages, fruit, and solid snacks show an overall increase, while yogurt remains relatively stable. For flavored tea, banana-flavored products exhibit minimal price changes, whereas strawberry-flavored products see a noticeable price increase. Interestingly, some banana-flavored products show significantly larger price increases compared to their strawberry counterparts, resulting in a more dispersed price distribution for banana-flavored items.

# Discussion {#sec-discussion}

## Summery of Findings

In our study, we analyzed the price changes of banana and strawberry-flavored products using both linear and Random Forest models. The linear model showed an average price change of 0.0091 for banana-flavored products and 0.0100 for strawberry-flavored products. The Random Forest model yielded similar results, with banana-flavored products averaging a price change of 0.0085, slightly lower than strawberry-flavored products at 0.0103. Both models indicated small price increases from July to November, with some differences in the distribution of price changes. July remained stable, while August and September saw slight declines. October and November experienced small increases, with the largest rise in October. Overall, price changes for both flavors stayed within ±0.1 units. Across food categories, beverages, fruits, and solid snacks showed price increases, while yogurt remained stable. Only in the flavored tea category did we observe a difference: banana-flavored products decreased, while strawberry-flavored ones increased. Overall, both models showed similar trends, with banana and strawberry products experiencing slight price increases, and strawberry-flavored products seeing a slightly higher increase.


This result surprised us, as our initial hypothesis assumed bananas to be more popular in winter and strawberries in summer. However, the model predicts that the price of strawberry-flavored products will increase more than that of banana-flavored ones. This suggests that while seasonal fruits may be more popular during their peak seasons, it does not necessarily mean their flavored products are less favored in off-seasons. For example, during winter, people might be even more inclined to purchase strawberry-flavored products to enjoy the taste of strawberries, rather than opting for banana-flavored ones.

Among all predictors, vendor and month are the most influential, while food category and average rainfall have relatively limited impact. Interestingly, the interaction between month and food category is the second most important predictor after month. We believe the dominance of vendor is due to the unique sales strategies each vendor employs, such as deciding which products to sell and how to adjust prices across different periods and categories. This makes vendor the most significant contributor to price changes, as vendors directly control pricing. The next most important predictors are month and its interaction with category, which reflects seasonal variations in consumer behavior and preferences for different food types. The interaction term provides deeper insights into these temporal consumption patterns. Although food category and average rainfall contribute less, both have importance values exceeding 20, demonstrating their limited but indispensable role in the model.

## Limitation

The model's subpar performance highlights a major limitation of this study: the quality of the original dataset, which significantly impacted the model's accuracy.

First, while data collection from vendor websites began on February 28, 2024, as noted by the dataset's creator, only a small number of products were captured until July 10, when large-scale scraping commenced. Unfortunately, strawberry- and banana-flavored products were not included in the earlier data. The earliest recorded prices for these products appear on June 11, leaving nearly four months of missing data. This substantial gap reduced both the dataset's size and its temporal coverage, making predictions more challenging.

Second, the dataset's diversity in product categories and inconsistent labeling across vendors resulted in significant data heterogeneity. For example, units of measurement were not standardized. Ideally, predictions should have been based on consistent units, such as price per 100g of bananas or per 100ml of strawberry beverages. However, some products were only listed with unit prices, while others had inconsistent units—for instance, yogurt prices were listed per 100g, per 100ml, or per box. This forced us to rely on unit prices, a less precise approach. For example, a 300ml beverage and a 2L beverage may have vastly different prices, but their per-100ml prices could be similar.

Third, the dataset had an uneven distribution of products between the two flavors. Strawberry-flavored products, likely due to their broader popularity or insufficient representation of banana-flavored products in the dataset, were far more prevalent. This resulted in denser data distributions for strawberry products compared to bananas, potentially introducing bias in banana product predictions. In some categories, such as flavored tea, the number of banana-flavored products was particularly sparse, further exacerbating prediction biases.

Lastly, our assumptions may not align with real-world dynamics. We hypothesized that seasonal fruits would see higher demand and prices during their peak seasons, with lower off-season demand. However, this may only hold for the fruits themselves and not their flavored products, which could be influenced by more significant factors than seasonality. Off-season sales, for instance, might not lead to price decreases but rather increases, as higher prices might offset lower demand. Overall, food price changes are influenced by numerous factors, making this a highly complex and nuanced subject. We underestimated the challenges of modeling price dynamics accurately.

## Future Study

Future research should address the limitations identified in this study to enhance the accuracy of modeling price changes for seasonal fruit-flavored products. First, improved data collection methods are essential. Future efforts should ensure consistent and comprehensive data coverage for all target product categories and flavors from the initial scraping period, ideally spanning 2–3 years to improve reliability. Second, standardizing product measurement units is crucial for enhancing prediction accuracy. Converting prices to uniform units, such as per 100 grams or per 100 milliliters, would reduce data heterogeneity and enable more meaningful comparisons. Efforts should also be made to unify units within the same food category and to resolve inconsistencies through manual data supplementation or improved web-scraping methods. For example, fruit prices could be collected uniformly per 100 grams, while beverages could be standardized per 100 milliliters. Products lacking detailed unit information, such as “boxed fruits” or “a case of strawberry juice,” could be supplemented with manual research or advanced scraping techniques. Lastly, incorporating additional predictive factors such as regional variations, promotional activities, or consumer preferences could provide deeper insights into the drivers of price changes. Leveraging these improvements, future studies could develop more robust models to better understand and forecast the dynamics of seasonal fruit-flavored product prices.

To build on the existing work with linear models and Random Forest, future studies could explore alternative machine learning algorithms that might better capture the complexities of flavored product price dynamics. Gradient Boosting Machines (GBMs), such as XGBoost or LightGBM, could be particularly effective, as they excel in handling structured data and capturing non-linear relationships while minimizing overfitting. Similarly, Support Vector Machines (SVMs) could be applied, especially if the focus shifts to classification tasks, such as identifying the likelihood of price increases or decreases based on features like flavor and category.


\newpage

# Appendix a. {-}

## Data Retrieval

We obtained the raw data from Jacob Filipp's Project Hammer website, available at: [https://jacobfilipp.com/hammer/](https://jacobfilipp.com/hammer/). To access the data, scroll down the page until you see the heading "Get Full Data in 2 CSV Files". Click on the link labeled "Zipped CSVs with full price data" to download the raw data. After extracting the files, you will obtain two CSV files: hammer-4-product.csv and hammer-4-raw.csv. Next, place these two files in the project's data/01-raw_data folder, and you’ll be able to run the code without issues.

(Note: When referencing file names and paths in R, use proper syntax with quotation marks, e.g., data/01-raw_data/hammer-4-product.csv.)

## Data Cleaning

The data cleaning process began with filtering the raw datasets, \texttt{hammer-4-raw.csv} and \texttt{hammer-4-product.csv}, to isolate and preprocess information about banana- and strawberry-related products. First, I removed rows where both \texttt{banana} and \texttt{strawberry} appeared in the \texttt{product\_name} or \texttt{concatted} columns to avoid mixed-category products. Subsequently, I filtered rows containing \texttt{banana} or \texttt{strawberry} in the \texttt{product\_name} column, creating separate datasets for banana (\texttt{product\_data\_banana}) and strawberry (\texttt{product\_data\_strawberry}) products. 

Next, I categorized products in each dataset based on keywords in the \texttt{product\_name} and \texttt{concatted} columns, assigning them to categories such as \texttt{Beverage}, \texttt{Yogurt}, \texttt{Flavored Tea}, \texttt{Solid Snacks}, or defaulting to \texttt{Fruit}. Unnecessary columns like \texttt{concatted}, \texttt{units}, \texttt{brand}, \texttt{detail\_url}, \texttt{sku}, and \texttt{upc} were then removed to streamline the data.

For the \texttt{raw\_data}, only rows with \texttt{product\_id}s present in the cleaned \texttt{product\_data\_banana} or \texttt{product\_data\_strawberry} datasets were retained. Rows with missing \texttt{price\_per\_unit} were removed, and the \texttt{nowtime} column was converted to a \texttt{Date} format. Additional columns, including \texttt{old\_price}, \texttt{price\_per\_unit}, and \texttt{other}, were removed. To handle missing dates, a sequence of dates for each \texttt{product\_id} was generated, and missing \texttt{current\_price} values were filled using the last known price.

Monthly average prices for each \texttt{product\_id} were calculated from June to November, and the results were merged back into \texttt{product\_data\_banana} and \texttt{product\_data\_strawberry}. Rows without any average price values across these months were removed from both datasets. Finally, the cleaned banana and strawberry datasets were saved as Parquet files in the \texttt{data/02-analysis\_data/} directory for further analysis.


# Appendix b. {-}

## Idealized Methodology for Analyzing Seasonal Price Changes of Fruit-Flavored Products

We aim to develop a comprehensive methodology for understanding price fluctuations of seasonal fruit-flavored products during their in-season and off-season periods. By combining consumer surveys and observational retail data, we will identify key trends in purchasing behavior and retailer pricing strategies. With a $100,000 budget, we employ stratified sampling, incentivized recruitment, and rigorous data validation to ensure robust and actionable insights. Results will be enriched through the integration of secondary data sources, including retail sales reports and online marketplace trends.

## Budget Allocation

Funding allocations are designed to ensure thorough data collection, effective sampling, and comprehensive analysis within a $100,000 budget. The proposed budget breakdown is as follows:

Survey platform costs	$15,000	Subscription fees for Qualtrics or Google Forms for dynamic and customizable surveys.

Respondent incentives	$20,000	Gift cards ($10 each) to encourage participation among 2,000 respondents.

Recruitment and staffing	$30,000	Staffing costs for survey distribution and observational data collection.

Observational data collection	$20,000	Travel and logistical costs for monitoring prices across various retail locations.

Data analysis tools	$10,000	Statistical software licenses and tools for data cleaning and analysis.

Marketing and promotion	$3,000	Social media campaigns to boost survey reach.
Contingency fund	$2,000	Reserved for unforeseen expenses.


## Sampling Approach

To ensure representation across demographics and consumer behavior, a stratified sampling method will be employed. The target population consists of regular consumers of fruit-flavored snacks, beverages, and desserts, focusing on age groups, income levels, geographic regions, and purchasing habits such as the frequency of buying seasonal fruit-flavored products. A sample size of 2,000 respondents has been selected, providing statistical robustness with a confidence level of 95% and a margin of error of ±2%. Recruitment efforts will primarily leverage online platforms, social media, and community networks to identify key focus groups, including frequent buyers who purchase products at least once a month, seasonal buyers who favor in-season purchases, and price-sensitive buyers whose decisions heavily depend on pricing.

Observational data collection will complement survey responses, focusing on retail locations, products, and time periods. A stratified sample of 30 retailers, encompassing supermarkets, convenience stores, and online marketplaces across urban, suburban, and rural areas, will be included. Observational metrics will focus on key fruit-flavored categories like strawberry, banana, and mixed-fruit products to capture price patterns. Data will be gathered over a one-year period to include full in-season and off-season cycles, offering a comprehensive view of pricing and availability trends.

## Respondent Recruitment and Data Validation

To obtain a representative and diverse dataset, a multi-mode recruitment strategy will be used. Online surveys distributed via platforms like Google Forms or Qualtrics will target food enthusiasts and frequent grocery shoppers through social media and email lists. We made a questionnnair, avaliable in following link: [https://forms.gle/gGy5CFKF33MCRm5ZA](https://forms.gle/gGy5CFKF33MCRm5ZA) In-store recruitment will involve flyers and QR codes displayed in participating retail stores, directly engaging shoppers. To encourage participation and ensure a high response rate, each respondent will be offered a $10 digital gift card upon survey completion. Data validation will include automated checks within the survey platform to flag incomplete or inconsistent responses and a manual review of open-ended answers to identify irrelevant or duplicate entries.

Observational data will provide real-world insights into price trends and product availability. Weekly price monitoring will record the prices of targeted products, while stock availability will be tracked to identify stockouts or limited supply during off-seasons. Observers will also document promotional campaigns and discounts to account for external influences on pricing.

## Analysis and Modeling

Survey and observational data will be integrated and analyzed using advanced statistical methods to uncover trends and insights. Weighted analysis will ensure that survey responses reflect the demographic composition of the target population. Observational data will be used for trend analysis, tracking seasonal price variations and comparing them across geographic regions. Predictive modeling, including regression analysis, will identify key factors influencing consumer willingness to pay for seasonal fruit-flavored products during in-season and off-season periods. Secondary data from retail sales reports and online platforms like Amazon and Walmart will further enrich the analysis, providing a comprehensive market perspective.

## Expected Outcomes

The study aims to generate actionable insights into consumer behavior and market trends. Consumer insights will highlight differences in purchasing behavior and price sensitivity during in-season and off-season periods, as well as the factors that influence willingness to pay for seasonal fruit-flavored products. Market trends will reveal seasonal pricing patterns, the impact of regional and demographic factors, and strategies employed by retailers to optimize pricing and promotions during off-seasons. These findings will offer valuable implications for businesses and policymakers seeking to align their strategies with consumer preferences.

\newpage

# References

